Simultaneous segmentation and distortion correction on diffusion weighted MR using shape priors

### 2012-12-18 Publications/2013-IPMI/Latex/02-Abstract.tex
Fixed some parts of latex

In whole-brain connectivity analysis of diffusion weighted MR images (DWI),
an accurate delineation of the white-matter and grey-matter surfaces is required.
While high-standard segmentation is readily available for anatomical MRI, such 
as T1-weighted, DWI typically have drastically lower resolution and strong 
geometrical distortions. We propose a DWI segmentation-registration framework 
that exploits the detailed anatomy extracted from anatomical MRI as shape-prior.
We use an ``active contours without edges''-like model to look for a deformation
field that optimally maps the shape prior on the multivariate features 
in diffusion space. This joint approach reflects the intrinsic coupling of 
segmentation and distortion correction. Complementary, a precise and consistent
cortical parcellation on DWI is straightforward by projection from T1 space. Thus,
we expect to improve the reliability and robustness of the resulting connectivity 
networks and their comparability within and across subjects. Preliminary results 
on synthetic datasets and simulated DWI confirm the effectiveness of our approach.

>>>>

In whole-brain connectivity analysis of diffusion weighted MR images (DWI),
an accurate delineation of the white-matter and grey-matter surfaces is required.
While high-standard segmentation is readily available for anatomical MRI, such 
as T1-weighted, DWI typically have drastically lower resolution and severe 
geometrical distortions. We propose a DWI segmentation-registration framework 
that exploits the detailed anatomy extracted from anatomical MRI as shape-prior.
We use an ``active contours without edges''-like model to look for a deformation
field that optimally maps the shape prior on the multivariate features 
in diffusion space. This joint approach reflects the intrinsic coupling of 
segmentation and distortion correction. Complementary, a precise and consistent
cortical parcellation on DWI is straightforward by projection from T1 space. Thus,
we expect to improve the reliability and robustness of the resulting connectivity 
networks and their comparability within and across subjects. Preliminary results 
on synthetic datasets and simulated DWI confirm the effectiveness of our approach.

### 2012-12-18 Publications/2013-IPMI/Latex/03-Introduction.tex
Fixed some parts of latex

In this work, we will refer as brain tissue segmentation to the precise
delineation of the \ac{csf}-\ac{gm} and \ac{gm}-\ac{wm} interface surfaces.
This segmentation is an important step on which strongly rely further
tasks. In tractography, a high-standard \ac{wm} mask is required. Otherwise,
there is an important risk for the algorithm to lose fiber bundles. This
requirement is usually solved in practice by plainly thresholding the 
\ac{fa} (a well-know scalar map derived from \ac{dwi} which depicts 
the isotropy of water diffusion inside the brain). 
Additionally, a precise location of the
\ac{gm}-\ac{wm} surface is required in the final steps to
achieve a consistent parcellisation of the cortex to represent the nodes 
of the output network. This parcellisation is generally defined in a 
high-resolution and better understood structural \ac{mri} of the same 
subject (e.g. T1 and/or T2 weighted acquisitions). Conversely, this 
problem is resolved with non-linear registration of a structural \ac{mri}
of the subject to the \ac{dwi} data. Even though some efforts have addressed
the study of the robustness of tractography versus the intra-subject variability
\cite{wakana_reproducibility_2007,heiervang_between_2006}, the results produced
are restricted to relevant regions of the brain. Therefore, extremely robust
and precise segmentation methods are required in the whole-brain application. \\

On the other hand, the \ac{dwi} data is usually obtained with \ac{epi}
acquisition techniques, that often suffer from severe distortions due to 
local field inhomogeneities. Generally, it is easily appreciated in the anterior
part of the brain, along the phase-encoded direction. Some methodologies have
been developed and generically named as \emph{\ac{epi}-unwarp} techniques
\cite{holland_efficient_2010,hsu_correction_2009,jezzard_characterization_2005,
reber_correction_2005}. They usually 
require the extra acquisition of the magnitude and phase of
the field (field-mapping), condition which is not always met. Some other 
methodologies do not make use the field-mapping, compensating the distortion
with non-linear registration from structural \ac{mri} or other means
\citep{andersson_modeling_2001}. To our knowledge, there exists no study
of the impact of the \ac{epi} distortion on the variability of tractography
results. \\

In this paper we propose a novel registration framework to simultaneously
solve the segmentation and distortion challenges, by exploiting as strong 
shape-prior the detailed anatomy extracted from anatomical \ac{mri}. This
is justified by the strong relationship between both problems, and the
advantage of a significant increase of coherence among these steps in
the complete processing pipeline.
We reformulate the segmentation problem as an inverse problem, where
we seek for an underlying deformation field (the distortion) mapping 
from the structural space into the diffusion space.

>>>>

In this work, we will refer as brain tissue segmentation to the precise
delineation of the \ac{csf}-\ac{gm} and \ac{gm}-\ac{wm} interface surfaces.
This segmentation is an important step on which strongly rely further
tasks. In tractography, a high-standard \ac{wm} mask is required. Otherwise,
there is an important risk for the algorithm to lose fiber bundles. This
requirement is usually solved in practice by plainly thresholding the 
\ac{fa}, a well-know scalar map derived from \ac{dwi} which depicts 
the isotropy of water diffusion inside the brain. 
Additionally, a precise location of the
\ac{gm}-\ac{wm} surface is required in the final steps to
achieve a consistent parcellisation of the cortex to represent the nodes 
of the output network. This parcellisation is generally defined in a 
high-resolution and better understood structural \ac{mri} of the same 
subject (e.g. T1 and/or T2 weighted acquisitions). Conversely, this 
problem is resolved with non-linear registration of a structural \ac{mri}
of the subject to the \ac{dwi} data. Even though some efforts have addressed
the study of the robustness of tractography versus the intra-subject variability
\cite{wakana_reproducibility_2007,heiervang_between_2006}, the results produced
are restricted to relevant regions of the brain. Therefore, extremely robust
and precise segmentation methods are required in the whole-brain application. \\

On the other hand, the \ac{dwi} data is usually obtained with \ac{epi}
acquisition techniques, that often suffer from severe distortions due to 
local field inhomogeneities. Generally, it is appreciated in the anterior
part of the brain, along the phase-encoded direction. Some methodologies have
been developed and generically named as \emph{\ac{epi}-unwarp} techniques
\cite{holland_efficient_2010,hsu_correction_2009,jezzard_characterization_2005,
reber_correction_2005}. They usually 
require the extra acquisition of the magnitude and phase of
the field (field-mapping), condition which is not always met. Some other 
methodologies do not make use the field-mapping, compensating the distortion
with non-linear registration from structural \ac{mri} or other means
\citep{andersson_modeling_2001}. To our knowledge, there exists no study
of the impact of the \ac{epi} distortion on the variability of tractography
results. \\

In this paper we propose a novel registration framework to simultaneously
solve the segmentation and distortion challenges, by exploiting as strong 
shape-prior the detailed anatomy extracted from anatomical \ac{mri}. On
one hand, the significant benefits of exploiting the anatomical \ac{mri}
have been demonstrated \cite{zollei_improved_2010}, justifying the use
of the shape prior information. On the other, the
significant coupling between both problems demands for consistent
segmentation, parcellation, and distortion. This need is satisfied by
the use of a unique underlying model of our joint approach.
We reformulate the segmentation problem as an inverse problem, where
we seek for an underlying deformation field (the distortion) mapping 
from the structural space into the diffusion space.

### 2012-12-18 Publications/2013-IPMI/Latex/045-Experiments.tex
Fixed some parts of latex

\section{Data and experiments}
\label{sec:experiments}
%
\subsection{Shape prior}
%
As described in \autoref{sec:introduction}, the general situation in
the connectivity pipelines consists on having 
a reliable segmentation obtained from the high resolution \ac{t1} 
reference image. Therefore, a precise location of the tissue interfaces
of interest is available in a reference space. Given that the anatomical reference segmentation is beyond the scope of this manuscript, we simply rely on the shape priors obtained from the models. \\

@@

The target \ac{dwi} data is characterized by its
low resolution (typically around $2.2x2.2x3mm^3$). Depending on the
posterior reconstruction methodology and the angular resolution
intended, the \ac{dwi} raw data has to be processed in order to
extract the information in a manageable manner. Here, we
will use the \ac{fa} and \ac{md} maps for convenience.
Whereas \ac{fa} describes the \emph{shape} of diffusion, 
the \ac{md} depicts the \emph{magnitude} of the process. 
There exist two main reasons to justify their choice. 
First, they are well-understood and standardized in clinical routine.
Second, together they contain most of the information that is
usually extracted from the \ac{dwi}-derived scalar maps. \\

For this model, we created manually a sound distortion visually similar
to real \ac{epi} distortions. We interpolated the distortion to a 
dense deformation field, necessary for warping the raw \ac{dwi} simulated
data. Once the signal was deformed, we proceeded to reconstruct the
\ac{dti} \textbf{FIXME is it really DTI, here, not rather DSI?} and subsequently obtained the scalars of interest (\ac{fa}, \ac{md}) and estimated their parameters on the model.

\textbf{FIXME Give parameters here?}\\

>>>>

\section{Data and experiments}
\label{sec:experiments}
%
\subsection{Shape prior}
%
As described in \autoref{sec:introduction}, the general situation in
the connectivity pipelines consists of having 
a reliable segmentation obtained from the high resolution \ac{t1} 
reference image. Therefore, a precise location of the tissue interfaces
of interest is available in a reference space. Given that the anatomical 
reference segmentation is beyond the scope of this manuscript, we simply 
rely on the shape priors obtained from the models.

@@

The target \ac{dwi} data is characterized by its distortions and its
low resolution (typically around $2.2x2.2x3mm^3$). Depending on the
posterior reconstruction methodology and the angular resolution
intended, the \ac{dwi} raw data has to be processed in order to
extract the information in a manageable manner. The properties of
the reconstructed tensors and derived scalar maps have been
studied and presented on \cite{ennis_orthogonal_2006}. Based on their
findings, the proposed energy model adapts to the \ac{fa} and \ac{md}
for their properties.
Whereas \ac{fa} describes the \emph{shape} of diffusion, 
the \ac{md} depicts the \emph{magnitude} of the process. 
There exist two main reasons to justify their choice. 
First, they are well-understood and standardized in clinical routine.
Second, together they contain most of the information that is
usually extracted from the \ac{dwi}-derived scalar maps. \\

For this model, we created manually a sound distortion visually similar
to real \ac{epi} distortions. We interpolated the distortion to a 
dense deformation field, necessary for warping the raw \ac{dwi} simulated
data. Once the signal was deformed, we proceeded to reconstruct the
\ac{dti} and subsequently obtained the scalars of interest (\ac{fa}, \ac{md}).
Finally, we estimated their parameters using the tissue probability
distribution maps from the original model (\autoref{table:parameters}).

\begin{table}
\begin{tabular}{cccc}
         & $\mathbf{\mu}_{FA}$ & $\mathbf{\mu}_{MD}$ & $\mathbf{\Sigma}$ \\
\ac{wm}  & & & \\
\ac{gm}  & & & \\
\ac{csf} & & & \\
\end{tabular}
\caption{Model parameters}
\label{table:parameters}
\end{table}
